---
title: "TCGA-BRCA Clinical Modeling and Survival Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TCGA-BRCA Clinical Modeling and Survival Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  warning   = FALSE,
  message   = FALSE,
  error     = FALSE,
  fig.width = 7,
  fig.height = 5
)

plot_or_message <- function(expr) {
  res <- try(eval.parent(substitute(expr)), silent = TRUE)

  if (inherits(res, "try-error")) {
    message("Plot call was evaluated but did not complete in this environment:")
    message(conditionMessage(attr(res, "condition")))
    return(invisible(NULL))
  }

  if (inherits(res, "ggplot")) {
    print(res)
  } else if (!is.null(res)) {
    print(res)
  }

  invisible(res)
}
```

## Overview

This vignette walks through the clinical modeling and survival-analysis tools in
**OmicsKit**, applied to the clinical features of the TCGA-BRCA (Breast Invasive
Carcinoma) cohort. Every model here is fit **live** on the clinical table so that
the code you read is the code that produced each figure.

The workflow mirrors a standard clinical-research pipeline:

1. **Exploratory survival profiling** — time-to-event curves with `nice_KM()`.
2. **Risk screening** — univariable hazard ratios with `get_cox()`.
3. **Risk summarization** — a forest plot of every model term with `nice_forest()`.
4. **Binary event prediction** — logistic models with `get_glm()`.
5. **Performance evaluation** — model discrimination with `nice_ROC()`.

## 1. Data Loading and Preparation

We load the package and the bundled clinical modeling table, then build a tidy
analysis frame. TCGA clinical fields use empty strings for missing values, so we
convert those to `NA` up front and keep only the columns we model.

```{r load-data}
library(OmicsKit)

data("brca_clinical_modeling_data")

model_vars <- c(
  "PFI", "PFI.time",
  "age_at_initial_pathologic_diagnosis",
  "AJCC_stage_broad", "histological_type",
  "menopause_status", "ER_group"
)

brca_model <- brca_clinical_modeling_data[, intersect(model_vars, names(brca_clinical_modeling_data))]

# Treat empty strings as missing so factor levels are clean.
brca_model[brca_model == ""] <- NA

# Collapse sparse categories so each level has enough events to model reliably.
brca_model$histology <- ifelse(
  brca_model$histological_type %in% c("Infiltrating Ductal Carcinoma",
                                      "Infiltrating Lobular Carcinoma"),
  brca_model$histological_type, "Other"
)
brca_model$histology[is.na(brca_model$histological_type)] <- NA

brca_model$menopause <- ifelse(grepl("^Pre",  brca_model$menopause_status), "Pre",
                        ifelse(grepl("^Post", brca_model$menopause_status), "Post", NA))
```

The clinical endpoint used throughout is the Progression-Free Interval (`PFI`,
event) and its follow-up time (`PFI.time`, in days):

```{r clinical-summary}
head(brca_model[, c("PFI.time", "PFI", "AJCC_stage_broad", "ER_group")])
```

---

## 2. Exploratory Survival Analysis: Kaplan-Meier Curves

`nice_KM()` builds Kaplan-Meier curves and annotates the Log-rank p-value. Here we
stratify the Progression-Free Interval by broad AJCC tumor stage. Note the
argument names required by the function: the grouping column goes in `gene`, and
the survival columns in `time_var` / `event_var`.

`nice_KM()` uses the grouping column's name (after `title_prefix`) as the legend
title, so we copy the stage into a cleanly named column — `Stage` — to keep the
legend free of underscores, and set `title_prefix = ""`.

```{r nice-km-example, fig.width = 8, fig.cap = "Kaplan-Meier PFI curves for TCGA-BRCA stratified by broad AJCC tumor stage."}
km_data <- brca_model[
  !is.na(brca_model$AJCC_stage_broad) &
    is.finite(brca_model$PFI.time) &
    !is.na(brca_model$PFI),
  ,
  drop = FALSE
]
km_data$Stage <- km_data$AJCC_stage_broad

plot_or_message(
  nice_KM(
    data         = km_data,
    gene         = "Stage",
    time_var     = "PFI.time",
    event_var    = "PFI",
    title_prefix = "",
    colors       = c("#2A9D8F", "#457B9D", "#E9C46A", "#E63946"),
    conf_int     = FALSE
  )
)
```

---

## 3. Risk Screening: Cox Proportional Hazards

As a first pass we screen each clinical covariate against progression risk with a
**univariable** Cox model per variable (`get_cox(model = "univariable")`). The
survival columns are passed via `time_col` / `event_col` and the predictors via
`vars`. Fitting one model per variable is robust to sparse categories — a
covariate that fails its event-count checks is skipped on its own instead of
taking the whole model down — and it yields a full set of terms to display.

```{r get-cox-example}
cox_table <- plot_or_message(
  get_cox(
    data      = brca_model,
    time_col  = "PFI.time",
    event_col = "PFI",
    vars      = c(
      "age_at_initial_pathologic_diagnosis",
      "AJCC_stage_broad",
      "ER_group",
      "histology",
      "menopause"
    ),
    model     = "univariable"
  )
)

cox_table[, c("variable", "term_clean", "reference", "HR", "CI_low", "CI_high", "p.value")]
```

---

## 4. Visualizing Hazard Ratios: Forest Plots

`nice_forest()` turns the tidy Cox table into a forest plot. By default it keeps
**every** model term (`p_display = 1`), so the plot shows both significant and
non-significant estimates — the point of a forest plot is to see the whole set of
effects and their confidence intervals side by side, not only the "winners".

Before plotting we do one cosmetic clean-up: the label columns carry
database-style names with underscores (`AJCC_stage_broad`, `ER_positive`,
`age_at_initial_pathologic_diagnosis`). A single `gsub()` step collapses any run
of underscores into one space across the three columns `nice_forest()` uses to
build its labels, so nothing on the plot shows an underscore:

```{r nice-forest-example, fig.width = 8, fig.height = 5.5, fig.cap = "Univariable clinical Cox screen for TCGA-BRCA PFI. Every model term is shown; colour marks statistical significance."}
# Replace single or repeated underscores with a single space for nicer labels.
label_cols <- c("variable", "term_clean", "reference")
cox_table[label_cols] <- lapply(cox_table[label_cols], function(x) gsub("_+", " ", x))

plot_or_message(
  nice_forest(
    data      = cox_table,
    title     = "TCGA-BRCA Univariable Clinical Cox Screen (PFI)",
    p_display = 1,
    sort_by   = "estimate"
  )
)
```

---

## 5. Binary Event Prediction: Logistic Regression

Alongside the time-to-event view, we can ask a **classification** question: from
baseline features alone, which patients go on to progress (`PFI = 1`) during
follow-up? `get_glm()` fits a logistic model and returns a tidy table of odds
ratios.

We use genuinely prognostic baseline predictors — tumor stage, age, receptor
status, histology, menopausal status — none of which are downstream of the
outcome, so there is no target leakage. (Treating `PFI` as a plain 0/1 label
ignores follow-up time; the Cox model in Section 3 is the time-aware analysis.
Here we use it only to demonstrate binary discrimination.)

```{r get-glm-example}
glm_result <- plot_or_message(
  get_glm(
    data       = brca_model,
    outcome    = "PFI",
    predictors = c(
      "AJCC_stage_broad",
      "age_at_initial_pathologic_diagnosis",
      "ER_group",
      "histology",
      "menopause"
    )
  )
)

glm_result
```

Tumor stage carries essentially all of the signal here — the odds of progression
rise steeply from Stage I through Stage IV.

---

## 6. Performance Evaluation: ROC Curves

To judge how well the baseline model separates progressors from non-progressors,
we evaluate it on a held-out test split and compare the full clinical model
against an age-only baseline. `nice_ROC()` takes a **named list of fitted `glm`
objects** plus the evaluation data and the binary `outcome`. `get_glm()` stores
the fitted model in `attr(result, "model")`, which we pass straight through.

```{r nice-roc-example, fig.cap = "ROC comparison for the progression classifier: full clinical model vs. age only."}
glm_frame <- brca_model[
  !is.na(brca_model$PFI),
  c("PFI", "AJCC_stage_broad", "age_at_initial_pathologic_diagnosis",
    "ER_group", "histology", "menopause")
]
glm_frame <- glm_frame[stats::complete.cases(glm_frame), , drop = FALSE]

set.seed(2025)
train_idx <- sample(nrow(glm_frame), size = floor(0.6 * nrow(glm_frame)))
train <- glm_frame[train_idx, , drop = FALSE]
test  <- glm_frame[-train_idx, , drop = FALSE]

fit_full <- attr(
  get_glm(
    data       = train,
    outcome    = "PFI",
    predictors = c("AJCC_stage_broad", "age_at_initial_pathologic_diagnosis",
                   "ER_group", "histology", "menopause")
  ),
  "model"
)

fit_age <- attr(
  get_glm(
    data       = train,
    outcome    = "PFI",
    predictors = "age_at_initial_pathologic_diagnosis"
  ),
  "model"
)

plot_or_message(
  nice_ROC(
    models     = list("Clinical model" = fit_full, "Age only" = fit_age),
    data       = test,
    outcome    = "PFI",
    plot_title = "Progression Classifier — ROC"
  )
)
```

---

## Session Info

```{r session-info}
sessionInfo()
```
