---
title: "TCGA-BRCA Omics-Layer Visualization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TCGA-BRCA Omics-Layer Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  warning   = FALSE,
  message   = FALSE,
  error     = FALSE,
  fig.width = 7,
  fig.height = 5
)

plot_or_message <- function(expr) {
  res <- try(eval.parent(substitute(expr)), silent = TRUE)

  if (inherits(res, "try-error")) {
    message("Plot call was evaluated but did not complete in this environment:")
    message(conditionMessage(attr(res, "condition")))
    return(invisible(NULL))
  }

  if (inherits(res, "ggplot")) {
    print(res)
  } else if (!is.null(res)) {
    print(res)
  }

  invisible(res)
}

include_optional_figure <- function(path) {
  if (file.exists(path)) {
    knitr::include_graphics(path)
  } else {
    cat("Figure not found: ", path, "\n", sep = "")
  }
}
```

## Overview

This vignette demonstrates how to visualize multi-omics data integrated into genomic coordinates using **OmicsKit**. This workflow focuses on:

1. Mapping and plotting omics features along linear chromosomes using `nice_GenomeTrack()`.
2. Generating whole-genome circular representations via `nice_circos()`.

## 1. Data Loading

We begin by loading the package and the multi-layer tracking dataset pre-computed for the TCGA-BRCA cohort.

```{r load-data}
library(OmicsKit)

# Load integrated genomic track coordinates and data
data("brca_omics_layer_tracks")
```

## 2. Linear Genomic Tracks (`nice_GenomeTrack`)

The `nice_GenomeTrack()` function allows you to plot specific molecular features or statistical metrics, such as log2 fold changes or mutation frequencies, across individual chromosomes or specific genomic regions.

This provides a clear one-dimensional overview of how molecular changes are spatially distributed across the genome.

```{r genome-track-call, fig.cap = "Linear genomic track visualization for selected chromosomes."}
plot_or_message(
  nice_GenomeTrack(
    tracks = brca_omics_layer_tracks$genome_tracks
  )
)
```

## 3. Circular Genome Visualization (`nice_circos`)

For a global, multi-layered perspective of the entire genome, a circular layout, commonly known as a Circos plot, is widely used in bioinformatics. The `nice_circos()` function integrates multiple tracks, such as copy number variation, expression levels, and clinical groups, into a single unified plot.

Since rendering interactive or highly detailed Circos plots during package compilation can be resource-intensive, the code block below is set to `eval = FALSE`. A pre-rendered high-resolution static image can be saved in the package figures directory for rapid documentation rendering.

```{r nice-circos-call, eval = FALSE}
# Example code to generate the full circular layout
nice_circos(
  tracks    = brca_omics_layer_tracks$circos_tracks,
  genome    = "hg38",
  sample_id = brca_omics_layer_tracks$sample_id
)
```

```{r circos-display, echo = FALSE, out.width = "90%", fig.cap = "Whole-genome circular multi-omics layout for TCGA-BRCA."}
include_optional_figure("figures/TCGA_A1_A0SH_genomewide_circos.jpg")
```

## Session Info

```{r session-info}
sessionInfo()
```
