---
title: "TCGA-BRCA Pathway Analysis and Clustering Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TCGA-BRCA Pathway Analysis and Clustering Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  warning   = FALSE,
  message   = FALSE,
  error     = FALSE,
  fig.width = 7,
  fig.height = 5
)

plot_or_message <- function(expr) {
  res <- try(eval.parent(substitute(expr)), silent = TRUE)

  if (inherits(res, "try-error")) {
    message("Plot call was evaluated but did not complete in this environment:")
    message(conditionMessage(attr(res, "condition")))
    return(invisible(NULL))
  }

  if (inherits(res, "ggplot")) {
    print(res)
  } else if (!is.null(res)) {
    print(res)
  }

  invisible(res)
}

include_optional_figure <- function(path) {
  if (file.exists(path)) {
    knitr::include_graphics(path)
  } else {
    cat("Figure not found: ", path, "\n", sep = "")
  }
}

```

## Overview

This vignette demonstrates a downstream workflow for Pathway Analysis (e.g., GSEA) using **OmicsKit**, applied to the TCGA-BRCA dataset. The workflow covers:

1. **Exploring enriched pathways** using single and multi-comparison visualizations (`splot_PA`, `multiplot_PA`).
2. **Extracting and annotating genes** from pathway results (`getgenesPA`, `addgenesPA`).
3. **Visualizing gene-level expression** inside selected pathways via heatmaps (`heatmap_PA`).
4. **Reducing redundancy** by computing pathway similarity and clustering pathway terms (`geneset_similarity`, `do_clust`).
5. **Detecting network communities** and identifying biological super-terms (`get_network_communities`, `network_clust_gg`).

Several computationally intensive visualizations are shown as pre-rendered static figures to keep vignette compilation fast and reproducible.

## 1. Data Loading

First, we load the package and the pre-computed TCGA-BRCA objects provided in the `data/` folder.

```{r load-data}
library(OmicsKit)

# Load pathway analysis results, gene set lists, and ranked genes
data("brca_pa_merged")
data("brca_geneset_list")
data("brca_ranked_genes")

# Load pre-computed clustering and similarity objects
data("brca_pa_similarity")
data("brca_pa_clustering")

```

Let's take a quick look at the merged pathway analysis results:

```{r overview-results}
table(brca_pa_merged$COMPARISON, brca_pa_merged$COLLECTION)

```

## 2. Visualization of Enriched Pathways

OmicsKit provides specialized functions to visualize GSEA results easily.

### Single Comparison (`splot_PA`)

We can visualize the significant pathways for a specific comparison, for example, *ER positive vs ER negative* within the Hallmark collection.

The code below shows how the plot was generated. The vignette displays a pre-rendered figure to avoid re-rendering a large pathway plot during package checks or website builds.

```{r splot-example-code, eval = FALSE}
pa_single <- subset(
  brca_pa_merged,
  COMPARISON == "ERpositive_vs_ERnegative" & COLLECTION == "Hallmark"
)

splot_PA(
  data           = pa_single,
  geneset_col    = "NAME",
  collection_col = "COLLECTION",
  nes_col        = "NES",
  fdr_col        = "FDR",
  order          = "desc",
  fill_limits    = c(0, 5),
  top_n_per_direction = 15,
  clean_labels   = TRUE,
  label_wrap     = 32,
  theme_params   = list(
    axis_title_size   = 12,
    axis_text_size_x  = 10,
    axis_text_size_y  = 7,
    strip_text_size   = 10,
    legend_title_size = 10,
    legend_text_size  = 9,
    bar_size          = 0.3,
    bar_width         = 0.72,
    hline_size        = 0.45
  )
)

```

```{r splot-example-display, echo = FALSE, out.width = "100%", fig.cap = "Top Hallmark pathway enrichments for the ER positive vs ER negative comparison."}
include_optional_figure("figures/pathway_splot_er_status.png")

```

### Multiple Comparisons (`multiplot_PA`)

If we want to compare enrichment patterns across multiple conditions, we can use `multiplot_PA()`. To keep this example compact, we focus on two comparisons and two estrogen-response Hallmark gene sets.

```{r multiplot-example-code, eval = FALSE}
pa_multi_subset <- subset(
  brca_pa_merged,
  COMPARISON %in% c("Tumor_vs_Normal", "ERpositive_vs_ERnegative") &
    NAME %in% c(
      "HALLMARK_ESTROGEN_RESPONSE_EARLY",
      "HALLMARK_ESTROGEN_RESPONSE_LATE"
    )
)

# Clean labels for display
pa_multi_subset$PATHWAY_LABEL <- pa_multi_subset$NAME
pa_multi_subset$PATHWAY_LABEL[
  pa_multi_subset$NAME == "HALLMARK_ESTROGEN_RESPONSE_EARLY"
] <- "Estrogen response\nearly"
pa_multi_subset$PATHWAY_LABEL[
  pa_multi_subset$NAME == "HALLMARK_ESTROGEN_RESPONSE_LATE"
] <- "Estrogen response\nlate"

comparison_labels <- c(
  "Tumor_vs_Normal" = "Tumor vs\nNormal",
  "ERpositive_vs_ERnegative" = "ER+ vs\nER-"
)

multiplot_PA(
  data             = pa_multi_subset,
  comparison_col   = "COMPARISON",
  facet_col        = "PATHWAY_LABEL",
  axis_y           = "NES",
  fdr_col          = "FDR",
  comparison_order = c("Tumor_vs_Normal", "ERpositive_vs_ERnegative"),
  custom_labels    = comparison_labels,
  ncol_wrap        = 2,
  free_y           = FALSE,
  fill_limits      = c(0, 5),
  theme_params     = list(
    axis_title_size     = 13,
    axis_text_size_x    = 10,
    axis_text_size_y    = 10,
    strip_text_size     = 11,
    hline_size          = 0.6,
    tick_size           = 0.4,
    tick_length         = 0.12,
    panel_spacing_multi = 0.8
  )
)

```

```{r multiplot-example-display, echo = FALSE, out.width = "100%", fig.cap = "Multi-comparison plot for selected Hallmark estrogen-response pathways."}
include_optional_figure("figures/pathway_multiplot_estrogen_response.png")

```

## 3. Gene Extraction and Annotation

Often, we need to know exactly *which* genes are associated with an enriched pathway result. Here, we extract genes for the same ER positive vs ER negative Hallmark subset.

```{r getgenes-pa-example}
pa_single <- subset(
  brca_pa_merged,
  COMPARISON == "ERpositive_vs_ERnegative" & COLLECTION == "Hallmark"
)

# Extract all pathway genes and leading-edge genes.
# The leading-edge option is lightweight and avoids requiring additional top-rank columns.
gene_lists <- plot_or_message(
  getgenesPA(
    pa_data      = pa_single,
    geneset_list = brca_geneset_list,
    ranked_genes = brca_ranked_genes$ERpositive_vs_ERnegative,
    genes        = c("all", "le")
  )
)

```

Once extracted, we can append this information directly into our results dataframe using `addgenesPA()`.

```{r addgenes-pa-example}
if (!is.null(gene_lists)) {
  pa_annotated <- addgenesPA(
    pa_data    = pa_single,
    gene_lists = gene_lists
  )

  # Check the new columns added
  print(head(pa_annotated[, c("NAME", "NES", "FDR", "le_genes")], 3))
}

```

## 4. Visualizing Gene Expression inside Pathways (Heatmaps)

OmicsKit's `heatmap_PA()` function automates the process of drawing expression heatmaps for genes associated with enriched pathways.

For this vignette, we display the leading-edge genes from `"HALLMARK_ESTROGEN_RESPONSE_LATE"` in the `"ERpositive_vs_ERnegative"` comparison. Samples are grouped by estrogen receptor status (`ER_group`) and ordered as ER Positive and ER Negative.

**Note:** The heatmap is displayed from a pre-rendered static image to keep vignette compilation lightweight and reproducible.

```{r heatmap-code, eval = FALSE}
# Example code to generate the heatmap
heatmap_PA(
  expression_data = brca_rna_vst_or_logexpr_small,
  metadata        = brca_rna_metadata_tumor_normal,
  pa_data_annot   = pa_annotated,
  ranked_genes    = brca_ranked_genes$ERpositive_vs_ERnegative,
  plot_genes      = "le_genes",
  sample_col      = "sampleID",
  group_col       = "ER_group",
  out_dir         = "figures/heatmaps"
)

```

```{r heatmap-display, echo = FALSE, out.width = "100%", fig.cap = "Leading-edge expression heatmap for HALLMARK_ESTROGEN_RESPONSE_LATE grouped by ER status."}
include_optional_figure("figures/heatmaps/HALLMARK_ESTROGEN_RESPONSE_LATE_heatmap.jpg")

```

## 5. Reducing Redundancy: Similarity and Clustering

Pathway analysis often returns redundant terms. OmicsKit reduces redundancy by computing pairwise pathway similarity and clustering related gene sets.

The code below illustrates how the silhouette plot can be inspected. The displayed figure is pre-rendered from the ER positive vs ER negative pathway similarity subset.

```{r similarity-clustering-code, eval = FALSE}
# Example code used during figure generation
brca_pa_clustering_er <- do_clust(brca_pa_similarity_er)
brca_pa_clustering_er$silhouette_plot

```

```{r similarity-clustering-display, echo = FALSE, out.width = "100%", fig.cap = "Silhouette plot used to evaluate pathway clustering for the ER positive vs ER negative subset."}
include_optional_figure("figures/pathway_silhouette_plot.png")

```

## 6. Network Communities and Super-terms

To assign biological meaning to pathway clusters, OmicsKit can detect communities within a pathway similarity network and extract representative biological super-terms.

The original similarity object contains multiple comparisons. For this vignette, the network figures were generated from the `ERpositive_vs_ERnegative` subset only. This keeps the examples focused on one biological contrast and avoids mixing nodes from multiple comparisons.

```{r networks-communities-code, eval = FALSE}
# Example code used during figure generation
comparison <- "ERpositive_vs_ERnegative"

brca_pa_similarity_er <- brca_pa_similarity
keep <- grepl(paste0("^", comparison, "::"), rownames(brca_pa_similarity$jaccard_sim))

brca_pa_similarity_er$jaccard_sim <- brca_pa_similarity$jaccard_sim[
  keep,
  keep,
  drop = FALSE
]

# Remove the comparison prefix from node names for cleaner labels
node_names <- sub(
  paste0("^", comparison, "::"),
  "",
  rownames(brca_pa_similarity_er$jaccard_sim)
)

rownames(brca_pa_similarity_er$jaccard_sim) <- make.unique(node_names)
colnames(brca_pa_similarity_er$jaccard_sim) <- make.unique(node_names)
brca_pa_similarity_er$dist_mat <- as.dist(1 - brca_pa_similarity_er$jaccard_sim)

brca_pa_clustering_er <- do_clust(brca_pa_similarity_er)

net_results_er <- get_network_communities(
  x             = brca_pa_similarity_er,
  threshold     = 0.3,
  method        = "louvain",
  superterms    = TRUE,
  n_terms       = 3,
  remove_prefix = TRUE,
  seed          = 174
)

head(net_results_er$superterms$summary)

network_clust_gg(
  x                 = brca_pa_similarity_er,
  clust_result      = brca_pa_clustering_er,
  jaccard_threshold = 0.3,
  min_degree        = 1,
  superterms        = TRUE,
  superterm_data    = net_results_er$superterms,
  type              = "superterms",
  seed              = 174
)

```

### Network without super-term labels

This version emphasizes the topology of the ER positive vs ER negative pathway similarity network without adding text labels.

```{r network-clean-display, echo = FALSE, out.width = "100%", fig.cap = "ER positive vs ER negative pathway similarity network without super-term labels."}
include_optional_figure("figures/pathway_network_er_clean_no_superterms.png")

```

### Network with community super-terms

This version labels pathway communities using automatically derived super-terms.

```{r network-superterms-display, echo = FALSE, out.width = "100%", fig.cap = "ER positive vs ER negative pathway similarity network with community-level super-term labels."}
include_optional_figure("figures/pathway_network_er_superterms.png")

```

### Individual community view

The individual view is useful when detailed community-level inspection is preferred over a single combined network representation.

```{r network-individual-display, echo = FALSE, out.width = "100%", fig.cap = "Individual ER positive vs ER negative pathway community view."}
include_optional_figure("figures/pathway_network_er_individual_with_superterms.png")

```

### Combined network representation

The combined view provides a compact summary of the ER positive vs ER negative pathway communities.

```{r network-combined-display, echo = FALSE, out.width = "100%", fig.cap = "Combined ER positive vs ER negative pathway community representation."}
include_optional_figure("figures/pathway_network_communities.png")

```

## Session Info

```{r session-info}
sessionInfo()

```
