---
title: "Pathway Analysis Clustering with OmicsKit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pathway Analysis Clustering with OmicsKit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5,
  warning   = FALSE,
  message   = FALSE
)
```

## Overview

After a gene set enrichment analysis (e.g., CAMERA, GSEA, fgsea), a common
challenge is that hundreds of significant gene sets are returned — many of which
are redundant because they share large overlapping gene memberships. This
vignette demonstrates how to use OmicsKit's pathway clustering functions to:

1. **Load gene sets** from GMT files with `list_gmts()`
2. **Quantify redundancy** between gene sets using Jaccard similarity with
   `calc_jaccard()`
3. **Cluster** redundant gene sets hierarchically with `do_clust()`
4. **Detect communities** in the gene set network and generate interpretable
   labels with `get_network_communities()`
5. **Visualize** the network with `network_clust()` (base R) or
   `network_clust_gg()` (ggplot2)

## Required packages

```{r packages}
library(OmicsKit)

# Suggested packages used in this vignette
library(igraph)
library(ComplexHeatmap)
```

---

## Step 1 — Load gene sets

In a real analysis, gene sets are stored as `.gmt` files. `list_gmts()` reads
all `.gmt` files in a directory and returns a single named list.

```{r list_gmts, eval = FALSE}
geneset_list <- list_gmts("path/to/your/gmt_folder/")
```

For this vignette we use the built-in example data, which contains 40 curated
gene sets spanning four biological themes: apoptosis, cell cycle, immune
response, and metabolism.

```{r load_data}
data(geneset_list)
data(camera_results)

# How many gene sets?
length(geneset_list)

# What does the enrichment results table look like?
head(camera_results)

# How many are significant at FDR < 0.05?
sum(camera_results$FDR < 0.05)
```

---

## Step 2 — Jaccard similarity matrix

`calc_jaccard()` filters the gene sets by FDR threshold and computes all
pairwise Jaccard similarity coefficients:

$$J(A, B) = \frac{|A \cap B|}{|A \cup B|}$$

A value of 1 means the two gene sets share identical gene membership; 0 means
no overlap at all.

```{r calc_jaccard}
jac <- calc_jaccard(
  geneset_list  = geneset_list,
  results       = camera_results,
  fdr_th        = 0.05
)

# The JaccardResult object contains three slots
names(jac)

# Dimensions of the similarity matrix
dim(jac$jaccard_sim)
```

The `$dist_mat` slot (1 − Jaccard similarity) can be reused independently —
for example as input to `nice_UMAP()` or any other distance-based method.

```{r jaccard_preview}
# Preview the top-left corner of the similarity matrix
jac$jaccard_sim[1:5, 1:5]
```

---

## Step 3 — Hierarchical clustering

`do_clust()` performs Ward-D2 hierarchical clustering on the distance matrix
and automatically selects the optimal number of clusters *k* by maximizing the
average silhouette width.

```{r do_clust}
clust <- do_clust(jac)

# Optimal k selected automatically
clust$optimal_k

# Cluster assignments (one row per gene set)
head(clust$cluster_assignments)
```

### Silhouette curve

The red dot marks the selected *k*:

```{r silhouette_plot, fig.cap = "Average silhouette width vs. number of clusters. The optimal k is highlighted in red."}
clust$silhouette_plot
```

### Jaccard heatmap with dendrogram

Gene sets in the same hierarchical cluster appear as blocks of high similarity
(darker blue) along the diagonal:

```{r heatmap, fig.width = 6, fig.height = 6, fig.cap = "Jaccard similarity heatmap with Ward-D2 dendrogram."}
ComplexHeatmap::draw(clust$heatmap)
```

---

## Step 4 — Community detection and super-terms

Hierarchical clustering groups gene sets by overall similarity, but the network
community detection captures a different structure: densely connected
sub-networks within the Jaccard graph.

`get_network_communities()` does three things in a single call:

- Builds a binary adjacency matrix (edges where Jaccard > `threshold`)
- Runs the Louvain algorithm to detect communities
- Calls `get_superterm()` internally to generate TF-IDF labels for each
  community

```{r get_network_communities}
net <- get_network_communities(
  x             = jac,
  threshold     = 0.3,
  method        = "louvain",
  superterms    = TRUE,
  n_terms       = 3,
  remove_prefix = TRUE,
  seed          = 1905
)

# How many communities were detected?
length(unique(net$membership))

# Community summary: label + size
net$superterms$summary
```

The `$superterms$mapping` tibble links every gene set to its community and
its super-term label:

```{r superterm_mapping}
head(net$superterms$mapping)
```

---

## Step 5 — Network visualization

### Option A: base R igraph (`network_clust`)

`network_clust()` draws plots directly to the active graphics device and
returns node attributes invisibly. It is faster and requires only `igraph`.

```{r network_clust_clean, fig.width = 7, fig.height = 7, fig.cap = "Gene set network — clean view. Node color reflects hierarchical cluster."}
result <- network_clust(
  x                 = jac,
  clust_result      = clust,
  jaccard_threshold = 0.10,
  min_degree        = 1,
  superterms        = TRUE,
  superterm_data    = net$superterms,
  type              = "clean",
  seed              = 1905
)
```

```{r network_clust_superterms, fig.width = 7, fig.height = 7, fig.cap = "Gene set network with community super-term labels."}
network_clust(
  x                 = jac,
  clust_result      = clust,
  jaccard_threshold = 0.10,
  min_degree        = 1,
  superterms        = TRUE,
  superterm_data    = net$superterms,
  type              = "superterms",
  seed              = 1905
)
```

### Option B: ggraph / ggplot2 (`network_clust_gg`)

`network_clust_gg()` returns a named list of ggplot2 objects that can be
further customized, saved with `ggsave()`, or composed with `patchwork`.
Requires `ggraph` and `tidygraph`.

```{r network_clust_gg, fig.width = 7, fig.height = 7, fig.cap = "Gene set network (ggraph) with super-term labels."}
plots <- network_clust_gg(
  x                 = jac,
  clust_result      = clust,
  jaccard_threshold = 0.10,
  min_degree        = 1,
  superterms        = TRUE,
  superterm_data    = net$superterms,
  type              = "all",
  seed              = 1905
)

# Display the combined view (super-terms + individual node labels)
plots$superterms
```

```{r patchwork, fig.width = 12, fig.height = 6, fig.cap = "Side-by-side comparison: clean view (left) vs. super-term labels (right)."}
library(patchwork)
plots$clean + plots$superterms
```

### Saving plots

```{r save, eval = FALSE}
# Save with ggsave
ggplot2::ggsave(
  filename = "network_combined.pdf",
  plot     = plots$combined,
  width    = 14,
  height   = 14
)

# Save base R igraph plot to PDF
pdf("network_superterms_igraph.pdf", width = 14, height = 14)
network_clust(jac, clust,
              superterms     = TRUE,
              superterm_data = net$superterms,
              type           = "superterms")
dev.off()
```

---

## Step 6 — Downstream tables

Both network functions return node-level attributes for downstream analysis:

```{r node_attributes}
# From the igraph version (invisible return)
head(result$node_attributes)
```

```{r superterm_report}
# Superterm report: community membership breakdown
result$superterm_report
```

---

## Full workflow — summary

```{r full_workflow, eval = FALSE}
library(OmicsKit)

# 1. Load gene sets
gsl <- list_gmts("path/to/gmt_folder/")

# 2. Jaccard similarity (filter by FDR < 0.05)
jac <- calc_jaccard(gsl, camera_results, fdr_th = 0.05)

# 3. Hierarchical clustering
clust <- do_clust(jac)
clust$silhouette_plot
ComplexHeatmap::draw(clust$heatmap)

# 4. Community detection + super-terms
net <- get_network_communities(jac, threshold = 0.3)
net$superterms$summary

# 5a. Network plot — base R (draws to device)
network_clust(jac, clust,
              superterms     = TRUE,
              superterm_data = net$superterms)

# 5b. Network plot — ggplot2 (returns objects)
plots <- network_clust_gg(jac, clust,
                          superterms     = TRUE,
                          superterm_data = net$superterms)
plots$combined
ggplot2::ggsave("network.pdf", plots$combined, width = 14, height = 14)

# 6. Reuse the distance matrix for UMAP
nice_UMAP(as.matrix(jac$dist_mat))
```

---

## Session info

```{r session_info}
sessionInfo()
```
